---
Title: Rotation 2 
Date: 2/4/2025
Objective: Look at the timing of regal mating behaviors from 1998 to 2015 using butterfly census data from FIG 
---
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

1. Sort data to only include regals that exhbit oviposition and mating behaviors (behaviors 6 and 8 respectively)
```{r}
regal_df<- updated_matrix |> 
  filter(butterfly_species== "ARID")


all_repro_df <- updated_matrix |> filter(Behavior %in% c(6,8))

regal_repro_df<- updated_matrix |> filter (butterfly_species =="ARID" & Behavior %in% c(6,8))

regal_repr_relabel <- regal_repro_df %>%
  mutate(behavior_type = case_when(
    Behavior == 6 ~ "ovipositing",
    Behavior == 8 ~ "mating",
    TRUE ~ "other"  # This covers any cases where Behavior is neither 6 nor 8
  ))

view (regal_repro_df)
```

2. Create a boxplot for each year with julian day on the x axis and number of butterflies with that behavior in the boxes
```{r}
regal_repro_df$Julian <- as.factor(regal_repro_df$Julian)
ggplot(regal_repro_df, aes(x = Julian, y = Year, fill = Behavior)) +
  geom_boxplot()# +
  facet_wrap(~ Year, scales = "free_y") +  # Creates separate panels for each year
  labs(title = "Behavior Timing Across Years", 
     x = "Behavior", 
      y = "Julian Day")+
  scale_x_continuous(labels = scales::number_format(accuracy = 1))

```

```{r}
behavior_count <- regal_repr_relabel |> 
  group_by(Year, Julian) |> 
  summarise(behavior_count= n(), .groups='drop')
ggplot(behavior_count, aes(x=Julian, y=Behavior_type))+
  geom_boxplot()+
  facet_wrap(~Year, scales= 'free_y')+
  scale_y_continuous(labels= scales::number_format(accuracy = 1))+
  theme(axis.text.x= element_text(angle =45, hjust=1))
```

```{r}
ggplot(all_repro_df, aes(x = Year, y = Julian, fill = Behavior)) +
  geom_boxplot() +
  labs(title = "Behavior Timing Across Julian Days by Sex", 
       x = "Behavior", 
       y = "Julian Day") +
  theme_minimal()
```

```{r}
all_behavior_count <- all_repro_df |> 
  group_by(Year, Julian) |> 
  summarise(behavior_count= n(), .groups='drop')
ggplot(all_behavior_count, aes(x=Julian, y=behavior_count))+
  geom_boxplot()+
  facet_wrap(~Year, scales= 'free_y')+
  scale_y_continuous(labels= scales::number_format(accuracy = 1))+
  theme(axis.text.x= element_text(angle =45, hjust=1))
```

Regression of Julian date of the observed behavior (response) by year (predictor) for each behavior
```{r}
install.packages("broom")
install.packages("purrr")
library(broom)
library (dplyr)
 #Step 1: Group data by 'Behavior'
# Step 2: Apply the regression for each behavior group
updated_matrix$Julian <- as.numeric(updated_matrix$Julian)
updated_matrix$Year <- as.numeric(updated_matrix$Year)
remove_na_df <- updated_matrix %>% drop_na(Julian, Year)
regression_results <- remove_na_df %>%
  group_by(Behavior) %>%
  group_split() %>%
  map(~ lm(Julian ~ Year, data = .)) %>%
  map_dfr(tidy)  # This extracts the coefficients and statistics

# View the regression results
print(regression_results)
```
```{r}
library(ggplot2)
ggplot(remove_na_df, aes(x = Year, y = Julian, color = factor(Behavior))) +
  geom_point(alpha = 0.6) +  # Scatter plot for the data points
  geom_smooth(method = "lm", se = FALSE, aes(group = Behavior), linetype = "solid") +  # Add regression lines
  labs(title = "Regression of Julian Day by Year for Each Behavior",
       x = "Year",
       y = "Julian Day",
       color = "Behavior") +
  theme_minimal() +
  theme(legend.position = "top")
```

Lineplot of all butterfly behavior frequency over Julian day (no year)
```{r}
behavior_freq <- updated_matrix %>%
  group_by(Julian, Behavior) %>%
  summarise(frequency = n()) %>%
  ungroup()

# Create the plot
ggplot(behavior_freq, aes(x = Julian, y = frequency, color = as.factor(Behavior))) +
  geom_line() +  # Line plot for frequencies
  labs(title = "Behavior Frequency by Julian Day",
       x = "Julian Day",
       y = "Frequency",
       color = "Behavior") +
  theme_minimal() +
  scale_color_discrete(name = "Behavior")
```
Line plot of regal behavior frequency by Julian (year not factored)
```{r}
regal_behavior_freq <- regal_df %>%
  group_by(Julian, Behavior, Year) %>%
  summarise(frequency = n()) %>%
  ungroup()


ggplot(regal_behavior_freq, aes(x = Julian, y = frequency, color = as.factor(Behavior), group = Behavior)) +
  geom_line(size = .75, alpha = 0.8) +  # Thicker lines with slight transparency
    # Add points for each behavior occurrence
  labs(title = "Behavior Frequency by Julian Day",
       x = "Julian Day",
       y = "Frequency",
       color = "Behavior") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +  # Use a distinct color palette for better visibility
  theme(legend.position = "right") +      # Move the legend to the right for clarity
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Whatever the hell 3D plot no one but Ian thought was a good idea
```{r}
library (plotly)

regal_behavior_freq$Behavior<- as.character(regal_behavior_freq$Behavior)
regal_behavior_freq$Year<- as.character(regal_behavior_freq$Year)
# Create the plotly 3D scatter plot
plot <- plot_ly(
  data = regal_behavior_freq,
  x = ~Julian,       # Julian Day on X axis
  y = ~Behavior,         # Behavior on Y axis
  z = ~frequency,
  color= ~Year, # Frequency on Z axis
  type = 'scatter3d',
  mode = 'lines' 
  # 3D scatter plot type
      # Plot markers
   # Customize marker size if needed
)
plot
```
Boxplot of all years by regal behaviors
```{r}
regal_df$Behavior<- as.character(regal_df$Behavior)
cut_behaviors<-regal_df |> filter(!Behavior %in% c(7,NA))
ggplot(cut_behaviors, aes(x=Julian, y=Behavior, fill = Behavior))+
  geom_boxplot()+
  facet_wrap(~Year, scales= 'free_y')+
  theme(axis.text.x= element_text(angle =45, hjust=1))
```
Boxplot of all butterfly behaviors by year
```{r}
unique(updated_matrix$Behavior)

updated_matrix$Year<- as.character(updated_matrix$Year)
all_cut_behaviors<-updated_matrix |> filter(!Behavior %in% c(7,NA))
ggplot(all_cut_behaviors, aes(x=Julian, y=Year, fill = Year))+
  geom_boxplot()+
  facet_wrap(~Behavior, scales= 'free_y')+
  theme(axis.text.x= element_text(angle =45, hjust=1))
```

Boxplot of all regal behaviors by year
```{r}
regal_df$Year<- as.character(regal_df$Year)
cut_behaviors<-regal_df |> filter(!Behavior %in% c(4,7,NA))
ggplot(cut_behaviors, aes(x=Julian, y=Year, fill = Year))+
  geom_boxplot()+
  facet_wrap(~Behavior, scales= 'free_y')+
  theme(axis.text.x= element_text(angle =45, hjust=1))
```

Distribution of flowers the regals are using
```{r}
regal_df |> filter

```

